Parsing Postfix log files

### 2009-03-05 doc/logparser-architecture.tex
Corrections to second half of architecture chapter.

Rules can specify conditions that will be evaluated by the framework before
attempting to use a rule to recognise an input: if the condition evaluates
to true the rule will be used, if not the rule will be skipped.  Conditions
can be as simple or complex as the parser requires, though naturally as the
complexity rises so does the difficulty in understanding how everything
interacts.  The framework has to evaluate each condition, so as the
complexity of conditions increases so does the complexity of the code
required to evaluate them; beyond a certain level of complexity conditions
should probably be written in a proper programming language, e.g.\ taking
advantage of dynamic languages' support for evaluating code at run-time, or
embedding a language like Lua.  If an implementation is going to use
conditions so complex that they will require a full programming language to
implement them, the design may need to be re-thought, including the choice
to use this architecture --- there may be other architectures more
suitable.

Conditions that examine the input will be the easiest to understand,
because they can be understood in isolation --- they do not depend on
variables set by actions or other rules.  Conditions that examine the input
may be complex, but simple conditions can be quite useful too, e.g.\ every
Postfix log \empty{}line contains the name of the Postfix component that
produced it, so every rule used in \parsername{} specifies the component
whose log \empty{}lines it recognises as a condition, reducing the number
of rules that will be used when classifying an input (see \sectionref{rules
in implementation} for details).

An example of how rule conditions can be useful is parsing C-style
comments, which start with \texttt{/*} and end with \texttt{*/}; the start
and end tokens can be on one line, or may have many lines between them.
\Tableref{Rules to parse C-style comments} shows the regexes, conditions,
and state changes of the four rules required to parse C-style comments.
These rules are somewhat simplified, e.g.\ rules one and two will recognise
the comment start token even if it is within a quoted string, which is
incorrect.  Rules one and two will only be used when the parser is parsing
code, not comments: rule one recognises a comment that is contained within
one line and leaves the parser's state unchanged; rule two recognises the
start of a command and changes the parser's state to parsing comments.
Rules three and four will be used when the parser is parsing a comment:
rule three recognises the end of a comment and switches parsing state back
to parsing code; rule four recognises a comment line without an end token
and keeps the parser's state unchanged.  It is important that the rules are
applied in the order listed in \tableref{Rules to parse C-style comments};
\sectionref{rules in architecture} has explained how this is achieved, and
\sectionref{overlapping rules in architecture} will discuss overlapping
rules.

\begin{table}[ht]
    \caption{Rules to parse C-style comments}
    \empty{}\label{Rules to parse C-style comments}
    \begin{tabular}{llll}
        \tabletopline{}%
        No.   & Regex             & Condition         & New State         \\
        \tablemiddleline{}%
        1     & \verb!/\*.*?\*/!  & state == code     & state = code      \\
        2     & \verb!/\*.*!      & state == code     & state = comment   \\
        3     & \verb!.*\*/!      & state == comment  & state = code      \\
        4     & \verb!.*!         & state == comment  & state = comment   \\
        \tablebottomline{}%
    \end{tabular}
\end{table}

Conditions can also check the value of variables that have been set by
either actions or rules; it will be easier to understand how a variable's
value will be used and changed if it is set by rules only rather than by
actions, because the chain of checking and setting can be followed from
rule to rule.  The downside to actions setting variables used in rule
conditions is action at a distance: to understand when a rule's condition
will be true requires understanding not just every rule but also every
action.  The framework probably does not need to support the same level of
complexity and flexibility when setting variables from rules (actions can
obviously use complex code if they are setting variables); however, if the
framework allows complex conditions, that code can probably be easily
extended to support complex variable assignments too.

@@

When adding new rules, the rule author must be aware that the new rule may
overlap with one or more existing rules, i.e.\ some inputs could be parsed
by more than one rule.  The architecture does not try to detect overlapping
rules: that responsibility is left to the author of the rules.
Unintentionally overlapping rules lead to inconsistent parsing and data
extraction because the first matching rule wins, and the order in which
rules are tried against each input might change between parser invocations.
Overlapping rules are frequently a requirement, allowing a more specific
rule to match some inputs and a more general rule to match the remainder,
e.g.\ separating \acronym{SMTP} delivery to specific sites from
\acronym{SMTP} delivery to the rest of the world.  Allowing overlapping
rules simplifies both the general rule and the more specific rule;
additionally rules from different sources can be combined with a minimum of
prior cooperation or modification required.  Overlapping rules should have
a priority attribute to specify their relative ordering; negative
priorities may be useful for catchall rules.

Decoupling the rules from the actions allows external tools to be written
to detect overlapping rules.  Traditional regexes are equivalent in
computational power to \acronym{FA} and can be converted to \acronym{FA},
so regex overlap can be detected by finding a non-empty intersection of two
\acronym{FA}\@.  The standard equation for \acronym{FA} intersection (given
for example in~\cite{intersection-of-NFA-using-Z}) uses DeMorgan's law:
$FA1~\cap{}~FA2~=~\overline{(\overline{FA1}~\cup{}~\overline{FA2})}$, which
has considerable computation complexity.  XXX COMMENT ON THE COMPUTATIONAL
COMPLEXITY OF INVERSION AND UNION\@; ADD A REFERENCE TO THE SPINNING WHEEL
BOOK\@.  Perl 5.10 regexes~\cite{perlre} are more powerful than traditional
regexes: it is possible to match correctly balanced brackets nested to an
arbitrary depth, e.g.\ \verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!
matches \verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires
the regex engine to maintain state on a stack, so Perl 5.10 regexes are
equivalent in computational power to \acronym{PDA}; detecting overlap may
require calculating the intersection of two \acronym{PDA} instead of two
\acronym{FA}.  The intersection of two \acronym{PDA} is not closed, i.e.\
the result cannot always be implemented using a \acronym{PDA}, so
intersection may be intractable in some cases, e.g.:
$a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
Detecting overlap amongst $n$ regexes requires calculating $n(n-1)/2$
intersections, resulting in $O(n^2x)$ complexity, where $O(x)$ is the
complexity of calculating \acronym{FA} or \acronym{PDA} intersection.  This
is certainly not a task to be performed every time the parser runs: naive
detection of overlap amongst \parsernames{} \numberOFrules{} rules would
require calculating \numberOFruleINTERSECTIONS{} intersections.  When
detecting overlap any conditions attached to rules must be taken into
account, because two rules whose regexes overlap may have conditions
attached which prevent the overlap in practice.  A less naive approach to
overlap detection would first check for overlapping conditions amongst
rules, and then check for overlapping regexes amongst the rules in each set
of rules with overlapping conditions.  Depending on the conditions employed
a rule may fall into more than one group, e.g.\ given these three
conditions:

@@

\noindent{}The first and second conditions do not overlap, but the third
condition does overlap with both the first and second conditions.  If the
rules are divided into sets based on how their conditions overlap, the
complexity of detecting overlap amongst $n$ rules is:

$$O(n^{2}y)~+~\sum{O(|s|^{2}x)~\forall{}~s~\in{}~overlaps}$$

@@

            $y$ & = & cost of checking for overlap between two conditions \\
            $x$ & = & cost of checking for overlap between two regexes    \\
     $overlaps$ & = & set of sets of rules with overlapping conditions    \\

@@

Once conditions pass a certain level of complexity determining whether two
conditions overlap becomes intractable, because it requires so much
knowledge of other rules and possibly even actions.  E.g.\ given two rules
with conditions \verb!verbose == true! and \verb!silent == true!, it
appears that these rules cannot overlap, yet there is nothing to stop both
variables being set to true by one or more other rules or actions.  If
variables used in conditions can be set by actions, determining whether two
conditions overlap is impossible, because understanding how an action will
change the variables is impossible; the Church-Turing Thesis XXX NEED A BIB
ENTRY FOR CHURCH-TURING explains why.

@@

It is possible to define pathological regexes which fall into two main
categories: regexes that match inputs they should not, and regexes that
consume excessive amounts of CPU time during matching.  Defining a regex
that matches inputs it should not is trivial: \verb!/^/! matches the start
of every input.  This regex would be found by a tool that detects
overlapping rules, and would easily be noticed by visual inspection, but
more complex regexes would be harder to spot.  Regexes that match more than
they should are a problem not because of excessive resource usage, but
because they may prevent the correct rule from recognising the input.  If
an adaptive ordering system is used (see \sectionref{rule ordering for
efficiency}) then a rule with a regex that matches too much may be promoted
up through the list, displacing an increasing number of correct rules as it
moves.

Usually excessive CPU time is consumed when a regex with a lot of
backtracking, caused by alteration and/or nested quantifiers, fails to
match; successful matching is generally quite fast with such regexes, so
problematic regexes may go unnoticed for some time.  E.g.\ with most regex
engines matching double quoted strings with \verb!"([^"\\]+|\\.)*"! is very
fast when there is a match, but when the match fails its computational
complexity is $O(2^{n})$; see~\cite{mastering-regular-expressions} for
in-depth discussion of nested quantifiers, backtracking, alteration, and
capturing groups.  Pathological regexes which consume excessive CPU time
can be difficult to detect, both by visual inspection and by machine
inspection, but if a regex is converted to a \acronym{FA} or the internal
representation used by the regex engine, it may be possible to determine if
there are nested quantifiers or other troublesome constructs present.
Modern regex engines have addressed many of these problems, e.g.\ the regex
to match double quoted strings given above fails immediately with Perl
5.10, regardless of the input length, because the regex engine looks for
the required double quotes first.  Similarly Perl 5.10's regex engine
optimises alterations starting with literal text into a trie, which has
constant matching time, whereas the matching time for alteration is
proportional to the number of alternatives.  Perl 5.10 regexes can use
\verb!(?>pattern)!, which matches \verb!pattern! the first time the regex
engine passes over it, and does not change what it originally matched if
the regex engine backtracks over it, so troublesome regexes can be modified
to use this to reduce the impact of backtracking; Prolog users will notice
a similarity to the \verb'!' (cut) operator.  A presentation showing some
of Perl 5.10's regex features is available at
\urlLastChecked{http://www.regex-engineer.org/slides/perl510_regex.html}{2009/03/03}.

Conditions can vary in complexity from simple equality through to a
Turing-equivalent language, so enumerating pathological conditions is
difficult if not pointless.  Conditions that examine the input using
regexes can suffer from the same problems as rule regexes; conditions that
check variables in uncomplicated ways may exhibit unexpected or incorrect
behaviour, but are unlikely to exhibit pathological behaviour.  Deciding
whether a more complex condition's behaviour is pathological or simply a
bug is difficult and to some extent is a matter of opinion.  The author has
not experimented with more complex conditions, so this thesis will not
address this point any further.

>>>>

Rules can have conditions attached that will be evaluated by the framework
before attempting to use a rule to recognise an input: if the condition is
true the rule will be used, if not the rule will be skipped.  Conditions
can be as simple or complex as the parser requires, though naturally as the
complexity rises so too does the difficulty in understanding how everything
interacts.  The framework has to evaluate each condition, so as the
complexity of conditions increases so will the complexity of the code
required to evaluate them.  Beyond a certain level of complexity,
conditions should probably be written in a proper programming language,
e.g.\ taking advantage of dynamic languages' support for evaluating code at
run-time, or embedding a language like Lua.  If an implementation is going
to use conditions so complex that they will require a Turing-equivilent
programming language to implement them, the design may need to be
re-thought, including the decision to use this architecture --- there may
be other architectures more suitable.

Conditions that examine the input will be the easiest to understand,
because they can be understood in isolation; they do not depend on
variables set by actions or other rules.  Conditions that examine the input
can be complex if required, but simple conditions can be quite useful too,
e.g.\ every Postfix log \empty{}line contains the name of the Postfix
component that produced it, so every rule used in \parsername{} has a
condition specifying the component whose log \empty{}lines it recognises,
reducing the number of rules that will be used when classifying an input
(see \sectionref{rules in implementation} for details) and increasing the
chance that the input will be correctly recognised..

Conditions can also check the value of variables that have been set by
either actions or rules; it is easier to understand how a variable's value
will be used and changed if it is set by rules only rather than by actions,
because the chain of checking and setting can be followed from rule to
rule.  The downside to actions setting variables used in rule conditions is
action at a distance: understanding when a rule's condition will be true or
false requires understanding not just every other rule but also every
action.  The framework probably does not need to support a high level of
complexity and flexibility when rules are setting variables; however, if
the framework supports complex conditions, that code can probably be easily
extended to support complex variable assignments too.  The level of
complexity the framework supports when evaluating conditions and setting
variables has two costs that must be taken into account when designing a
parser: the difficulty of implementation, and the difficulty of
understanding rules and writing correct rules.  XXX NEEDS TO BE IMPROVED\@.

An example of how rule conditions can be useful is parsing C-style
comments, which start with \texttt{/*} and end with \texttt{*/}; the start
and end tokens can be on one line, or may have many lines between them.
\Tableref{Rules to parse C-style comments} shows the regexes, conditions,
and state changes of the four rules required to parse C-style comments.
These rules are somewhat simplified, e.g.\ rules one and two will recognise
the comment start token even if it is within a quoted string, which is
incorrect.  Rules one and two will be used when the parser is parsing code,
not comments: rule one recognises a comment that is contained within one
line and leaves the parser's state unchanged; rule two recognises the start
of a comment and changes the parser's state to parsing comments.  Rules
three and four will be used when the parser is parsing a comment: rule
three recognises the end of a comment and switches the parser's state back
to parsing code; rule four recognises a comment line without an end token
and keeps the parser's state unchanged.  It is important that the rules are
applied in the order listed in \tableref{Rules to parse C-style comments};
\sectionref{rules in architecture} has explained how this is achieved, and
\sectionref{overlapping rules in architecture} will discuss overlapping
rules.

\begin{table}[ht]
    \caption{Rules to parse C-style comments}
    \empty{}\label{Rules to parse C-style comments}
    \begin{tabular}{llll}
        \tabletopline{}%
        No.   & Regex             & Condition           & Variable Changes    \\
        \tablemiddleline{}%
        % XXX WHAT SHOULD THE VARIABLE BE CALLED?
        1     & \verb!/\*.*?\*/!  & parsing == code     & parsing = code      \\
        2     & \verb!/\*.*!      & parsing == code     & parsing = comment   \\
        3     & \verb!.*\*/!      & parsing == comment  & parsing = code      \\
        4     & \verb!.*!         & parsing == comment  & parsing = comment   \\
        \tablebottomline{}%
    \end{tabular}
\end{table}

@@

When adding new rules, the rule author must be aware that the new rule may
overlap with one or more existing rules, i.e.\ some inputs could be parsed
by more than one rule.  Unintentionally overlapping rules lead to
inconsistent parsing and data extraction because the first matching rule
wins, and the order in which rules are tried against each input might
change between parser invocations.  Overlapping rules are frequently a
requirement, allowing a more specific rule to match some inputs and a more
general rule to match the remainder, e.g.\ separating \acronym{SMTP}
delivery to specific sites from \acronym{SMTP} delivery to the rest of the
world.  Allowing overlapping rules simplifies both the general rule and the
more specific rule; additionally rules from different sources can be
combined with a minimum of prior cooperation or modification required.
Overlapping rules should have a priority attribute to specify their
relative ordering; negative priorities may be useful for catchall rules.
The architecture does not try to detect overlapping rules: that
responsibility is left to the author of the rules.

Overlapping rules can be detected by visual inspection, or a program may be
written to analyse the rule's regexes.  Traditional regexes are equivalent
in computational power to \acronym{FA} and can be converted to
\acronym{FA}, so regex overlap can be detected by finding a non-empty
intersection of two \acronym{FA}\@.  Perl 5.10 regexes~\cite{perlre} are
more powerful than traditional regexes: it is possible to match correctly
balanced brackets nested to an arbitrary depth, e.g.\
\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!  matches
\verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires the
regex engine to maintain state on a stack, so Perl 5.10 regexes are
equivalent in computational power to \acronym{PDA}; detecting overlap may
require calculating the intersection of two \acronym{PDA} instead of two
\acronym{FA}.  \acronym{PDA} intersection is not closed, i.e.\ the result
cannot always be implemented using a \acronym{PDA}, so intersection may be
intractable in some cases, e.g.:
$a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
Detecting overlap amongst $n$ regexes requires calculating $n(n-1)/2$
intersections, resulting in $O(n^2x)$ complexity, where $O(x)$ is the
complexity of calculating \acronym{FA} or \acronym{PDA} intersection.  This
is certainly not a task to be performed every time a parser runs: naive
detection of overlap amongst \parsernames{} \numberOFrules{} rules would
require calculating \numberOFruleINTERSECTIONS{} intersections.

When detecting overlap any conditions attached to rules must be taken into
account, because two rules whose regexes overlap may have conditions
attached which prevent the rules overlapping.  A less naive approach to
overlap detection would first check for overlapping conditions amongst
rules, and then check for overlap between the regexes of each pair of rules
with overlapping conditions.  Depending on the conditions employed a rule
may overlap with multiple rules, and rule overlap is not transitive, e.g.\
given these three conditions:

@@

XXX CARL SUGGESTS COMPARING THIS TO THE POWERSET\@; IS THAT REASONABLE\@?

\noindent{}The first and second conditions do not overlap, but the third
condition overlaps with both the first and second conditions.  When rules
are divided into sets based on how their conditions overlap, the complexity
of detecting overlap amongst $n$ rules is:

XXX IN THE FORMULA BELOW $s \in{} overlaps$ SHOULD BE UNDER $\sum$

$$O(n^{2}y) + O\left(\sum{|s|^{2}x \forall{} s \in{} overlaps}\right)$$

@@

            $x$ & = & cost of checking for overlap between two regexes    \\
            $y$ & = & cost of checking for overlap between two conditions \\
     $overlaps$ & = & set of sets of rules with overlapping conditions    \\

@@

For this approach to be more efficient than the naive approach $y$ must be
significantly lower than $x$.  If $y$ is higher than $x$ then the opposite
approach should be taken: check for regex overlap first, then check
for condition overlap only between rules that have overlapping regexes.

Once conditions pass a certain level of complexity determining whether two
conditions overlap becomes intractable, because it requires so much
knowledge of other rules and possibly even actions.  E.g.\ given two rules
with conditions \verb!verbose == true! and \verb!silent == true!, logically
these rules should not overlap, yet there is nothing to stop both variables
being set to true by one or more rules or actions.  If variables used in
conditions can be set by actions, determining whether two conditions
overlap is impossible, because understanding how an action will change the
variables is impossible.  XXX DO I NEED TO MENTION CHURCH-TURING OR
ANYTHING\@?

@@

It is possible to define pathological regexes which fall into two main
categories: regexes that match inputs they should not, and regexes that
consume excessive amounts of CPU time during matching.  Defining a regex
that matches inputs it should not is trivial: \verb!/^/! matches the start
of every input.  This regex would be found by a tool that detects
overlapping rules, and would easily be noticed by visual inspection, but
more complex regexes would be harder to spot.  Regexes that match more than
they should are a problem not because of excessive resource usage, but
because they may prevent the correct rule from recognising the input.  If
an adaptive ordering system is used that prioritises rules that frequently
recognise inputs (see \sectionref{rule ordering for efficiency}), then a
rule with a regex that matches inputs it should not may be promoted up
through the list, displacing an increasing number of correct rules as it
moves.

Usually excessive CPU time is consumed when a regex with a lot of
backtracking, caused by alteration and/or nested quantifiers, fails to
match; successful matching is generally quite fast with such regexes, so
problematic regexes may go unnoticed for some time.  E.g.\ with most regex
engines matching double quoted strings with \verb!"([^"\\]+|\\.)*"! is very
fast when there is a match, but when the match fails its computational
complexity is $O(2^{n})$; see~\cite{mastering-regular-expressions} for
in-depth discussion of nested quantifiers, backtracking, alteration, and
capturing groups.  Pathological regexes which consume excessive CPU time
can be difficult to detect, both by visual inspection and by machine
inspection, but if a regex is converted to a \acronym{FA} or the internal
representation used by the regex engine, it may be possible to determine if
there are nested quantifiers or other troublesome constructs present.
Modern regex engines have addressed many of these problems, e.g.\ the regex
to match double quoted strings given above fails immediately with Perl
5.10, regardless of the input length, because the regex engine looks for
both of the required double quotes first.  Similarly Perl 5.10's regex
engine optimises alterations starting with literal text into a trie, which
has matching time proportional to the length of the alternatives, rather
than the number of alternatives.  Perl regexes can use \verb!(?>pattern)!,
which matches \verb!pattern! the first time the regex engine passes over
it, and does not change what it originally matched if the regex engine
backtracks over it, so troublesome regexes can use this to reduce the
impact of backtracking; Prolog users will notice a similarity to the
\verb'!' (cut) operator.  A presentation showing some of Perl 5.10's new
regex features is available at
\urlLastChecked{http://www.regex-engineer.org/slides/perl510_regex.html}{2009/03/03}.

Conditions can vary in complexity from simple equality through to a
Turing-equivalent language, so enumerating pathological conditions is
difficult if not pointless.  Conditions that examine the input using
regexes can suffer from the same problems as rule regexes; conditions that
check variables in uncomplicated ways may exhibit unexpected or incorrect
behaviour, but are unlikely to exhibit pathological behaviour.  Deciding
whether a more complex condition's behaviour is pathological or simply a
bug is difficult and to some extent is a matter of opinion.  When this
architecture has received more widespread usage consensus should be reached
on the topic of pathological conditions.
